---
title: "Carbon fixation"
author: "george.westmeijer@umu.se"
date: now
format: html
toc: true
number_sections: true
theme: sandstone
mainfont: Helvetica
monofont: AndaleMono
fontsize: "8"
dpi: 300
highlight: textmate
df_print: paged
code-fold: true
code-summary: "Expand code"
self_contained: true
embed-resources: true
keep_md: false
encoding: "UTF-8"
---

```{r}
#| label: setup
#| include: false
#| echo: false

knitr::opts_chunk$set(cache = TRUE)
```


```{r}
#| label: libraries
#| cache: false
#| include: false

library(dplyr)
library(tidyr)
library(tibble)
library(forcats)
library(ggplot2)
library(readr)
library(patchwork)
library(grateful)
library(ggtree)
library(mclink)
```

# Read input files

The file 'c-fixation-kos.gtdb-hits.tsv.gz' contains the genomes' accessions that have one or more hits of KOs part of a carbon fixation module.  


```{r}
#| label: read-taxonomy

gtdb <- read_tsv('../gtdb/ar53_metadata_r232.tsv.gz', show_col_types = F) %>%
  select(accession, ncbi_genbank_assembly_accession, taxonomy = gtdb_taxonomy, gtdb_representative) %>%
  mutate(accession = gsub('^.._', '', accession)) %>%
  mutate(taxonomy = gsub('.__', '', taxonomy))

gtdb <- read_tsv('../gtdb/bac120_metadata_r232.tsv.gz', show_col_types = F) %>%
  select(accession, ncbi_genbank_assembly_accession, taxonomy = gtdb_taxonomy, gtdb_representative) %>%
  mutate(accession = gsub('^.._', '', accession)) %>%
  mutate(taxonomy = gsub('.__', '', taxonomy)) %>%
  rbind(gtdb) %>% 
  pivot_longer(1:2, values_to = 'accession') %>% 
  select(accession, taxonomy, gtdb_representative) %>% 
  separate(taxonomy, into = c('domain', 'phylum', 'class', 'order', 'family', 'genus', 'species'), sep = ';') %>% 
  distinct() %>% 
  # Keep only species representatives
  filter(gtdb_representative) %>% 
  group_by(phylum) %>% mutate(n.rep = n()) %>% ungroup() %>% 
  select(accession, domain, phylum, n.rep) %>% distinct()
```


```{r}
#| label: read-kofam

hits <- read_tsv('data/c-fixation-kos.gtdb-hits.tsv.gz', col_types = 'ccccdddc') %>%
  select(ko, accession = genome_accno) %>%
  distinct() %>%
  mutate(presence = 1) %>% 
  pivot_wider(names_from = 'accession', values_from = 'presence', values_fill = 0)

markers <- read_tsv('data/ko_fixation.tsv', col_types = 'cccccclcl') %>% 
  select(ko, module, marker) %>% 
  filter(marker) %>% 
  mutate(module = gsub('https://www.kegg.jp/module/', '', module)) %>% 
  select(-marker)
```


```{r}
#| label: mclink

# M00165 (CBB) has 12 blocks
# M00173 (Reverse citrate) has <nightmare>
# M00374 (DC-HB cycle) has 21 blocks
# M00375 (HP/HB cycle) has 17 blocks
# M00376 (3-HP bi-cycle) has 16 blocks (15 in the case of K01847)
# M00377 (WL pathway) has 14 blocks (13 in the case of K05299 + K15022)

# This is the actual analysis
mcl <- KO_pathway_ref %>% 
  filter(Module_Entry %in% c('M00165','M00377','M00173','M00376','M00375','M00374')) %>% 
  mclink(data = data.frame(hits), ref = .)

# This extracts the data
scores <- mcl[['coverage']] %>% 
  as_tibble() %>% 
  rename(module = Module_Entry) %>%
  select(-Level_2, -Level_3, -Definition, -Module_Name) %>% 
  pivot_longer(2:ncol(.), names_to = 'accession', values_to = 'score') %>% 
  left_join(gtdb, by = 'accession') 
```

```{r}
#| label: score-cutoff

cutoff <- scores %>% 
 # If the module score < 0.75, count it as complete
  mutate(cutoff = if_else(score > 0.75, 1, 0)) %>% 
  # Average the scores for each phylum/module combination
  group_by(phylum, module) %>% 
  # Divide the number of genes with a full module ('cutoff') by the number of genomes in that phylum ('n.rep')
  summarise(
    score = sum(cutoff),
    .groups = 'drop') %>%
  mutate(module = case_when(
    module == 'M00165' ~ 'CBB cycle',
    module == 'M00377' ~ 'Wood-Ljungdahl',
    module == 'M00173' ~ 'Reverse TCA',
    module == 'M00376' ~ '3-HP bi-cycle',
    module == 'M00375' ~ 'HP-HB cycle',
    module == 'M00374' ~ 'DC-HB cycle'
  ))
```



# KO matrix

The mclink function is used to calculate the completeness of the six relevant modules (M00165, M00377, M00173, M00376, M00375, M00374). The presence of any KO for each step indicates completeness, and the completeness is a score ranging from 0 to 1. The average value of this score is the genome-specific module completeness score. For example, a score of 0.9 for a module containing 10 steps (reactions) implies 9/10 genes were present.  


# Archaea
## Collapse archaeal phylogeny on phylum level


```{r}
#| label: archaeal-representatives

# Read the tree (archaeal) tree and remove the 'GB' or 'RS' prefix
tree <- ape::read.tree('../gtdb/ar53_r232.nwk')
tree$tip.label <- gsub('^.._', '', tree$tip.label)

# getMRCA() only works when two or more tips are provided. So remove clades with a single representative

i <- gtdb %>%
  filter(domain == 'Archaea') %>% 
  select(accession, phylum) %>% 
  # Consider only genomes represented in the tree
  filter(accession %in% tree$tip.label) %>%
  tibble::deframe()

# Assigns the accessions of each phyla in a list
groups <- split(names(i), i)

# Compute one representative node per phylum
phy_nodes <- sapply(groups, function(tips) {
  if (length(tips) > 1) {
    ape::getMRCA(tree, tips) # If the clade has more than one genome, find the MRCA
  } else {
    which(tree$tip.label == tips) # If only one tip, return its position in the tree
  }
})
```




```{r}
#| label: archaeal-nodes

# Ensure node indices are numeric and named by phylum (if all is well, this shouldn't change the data)
phy_nodes <- as.numeric(phy_nodes)
names(phy_nodes) <- names(groups)

#node <- phy_nodes[[fam]]
#tree$node.label <- rep(NA, tree$Nnode) # Remove node labels and replace with NA

# Indices for tip labels
tip_idx <- seq_along(tree$tip.label)

# Assign phylum names either to tips or internal nodes
for (p in names(phy_nodes)) {
  node <- phy_nodes[[p]]
  
  if (node <= length(tree$tip.label)) {
    # If node index corresponds to a tip - rename the tip
    tree$tip.label[node] <- p
  } else {
    # Otherwise it's an internal node → assign node label
    tree$node.label[node - length(tree$tip.label)] <- p
  }
}
```




```{r}
#| label: collapse-archaeal-tree

arctree <- ggtree(tree, layout = 'circular', branch.length = 'none') # This reads the original phylogeny into a ggtree object

# Collapse clades (multi-tip phyla) into single branches
for (p in names(groups)) {
  tips <- groups[[p]]
  
  if (length(tips) > 1) {
    node <- phy_nodes[p]
    if (!is.null(node)) {
      arctree <- collapse(arctree, node = node)
    }
  }
}
```




```{r}
#| label: highlight-archaeal-nodes

for (p in names(phy_nodes)) {
  node <- phy_nodes[p]
  
  if (!is.null(node)) {
    arctree <- arctree + geom_hilight(node = node, alpha = 0.3)
  }
}
```






```{r}
#| label: plot-archaea

arctree +
  geom_tiplab(size = 2, align = TRUE) +
  geom_nodelab(aes(label = label), size = 2, hjust = 0)

```

```{r}
#| label: export-archaeal-phyla

ggsave(filename = 'figures/gtdb_r232_arc_phylum.png', width = 180, height = 240, units = 'mm')
```



## Collapse archaeal tree with single representative

```{r}
#| label: drop-nwk

# There are 25 archaeal phyla in GTDB 232

tree <- ape::read.tree('../gtdb/ar53_r232.nwk')
tree$tip.label <- gsub('^.._', '', tree$tip.label) # 25 archaeal phyla in the tree

# Select a single representative per phylum, 25 in total
rep <- gtdb %>%
  filter(domain == 'Archaea' & gtdb_representative) %>% 
  group_by(phylum) %>%
  slice(1) %>% pull(accession)

# Drop the genomes not among the representatives
tree <- ape::keep.tip(tree, rep)

# Decorate the tip labels with phyla names
tree$tip.label <- tree$tip.label %>% 
  tibble() %>% 
  rename(accession = 1) %>% 
  left_join(gtdb, by = 'accession') %>% 
  pull(phylum)

# Export the tree
ape::write.tree(tree, file = 'data/arc_phyla_226.nwk')
```


```{r}
#| label: hits-score

scores <- tree$tip.label %>% 
  tibble() %>% select(phylum = 1) %>% 
  left_join(cutoff, by = 'phylum', relationship = 'one-to-many') %>% 
  replace_na(list('module' = 'CBB cycle', 'score' = 0)) %>% 
  # Some phyla are not among the 'hits', thus have a score of 0
  pivot_wider(names_from = 'module', values_from = 'score', values_fill = 0) %>% 
  mutate(phylum = factor(phylum, levels = tree$tip.label)) %>% 
  column_to_rownames('phylum')
```


```{r}
#| label: complete-pathway-archaea

complete <- kos %>%
  mutate(threshold = if_else(score > 0.75, 1, 0)) %>% 
  # Number of genomes with marker gene or complete pathway per phylum
  group_by(phylum) %>% 
  summarise(n = n() / 6, n_complete = sum(threshold), .groups = 'drop') %>% 
  mutate(tip.lab = paste(phylum, ' (', n_complete, '/', n, ')', sep = ''))
```


```{r}
#| label: marker-genes-archaea

marker.completeness <- hits %>% 
  pivot_longer(-1, names_to = 'accession', values_to = 'encoded') %>% 
  filter(encoded > 0) %>% 
  select(-encoded) %>% 
  inner_join(markers, by = 'ko', relationship = 'many-to-many') %>% 
  # Check if the multiple subunits of the marker gene are present
  group_by(accession, module) %>% 
  summarise(n = n(), .groups = 'drop') %>% 
  mutate(complete = case_when(
    module == 'M00165' & n == 2 ~ TRUE,
    module == 'M00173' & n == 7 ~ TRUE,
    module == 'M00374' & n == 1 ~ TRUE,
    module == 'M00375' & n == 1 ~ TRUE,
    module == 'M00376' & n == 4 ~ TRUE,
    module == 'M00377' & n == 1 ~ TRUE,
    .default = FALSE
  )) %>% 
  filter(complete) %>% 
  left_join(gtdb, by = 'accession') %>%
  # Summarise per phylum
  group_by(phylum, module) %>% summarise(n.marker = n(), .groups = 'drop') %>% 
  left_join(n.arc, by = 'phylum', relationship = 'many-to-one')

i <- tree$tip.label %>% 
  tibble() %>% select(phylum = 1) %>% 
  left_join(marker.completeness, by = 'phylum', relationship = 'one-to-many') %>%
  group_by(phylum, module) %>% summarise(p.marker = n.marker / n, .groups = 'drop') %>% 
  #mutate(p.marker = n.marker / n) %>% 
  replace_na(list('module' = 'M00165', 'p.marker' = 0)) %>% 
  # Some phyla are not among the 'hits', thus have a score of 0
  pivot_wider(names_from = 'module', values_from = 'p.marker', values_fill = 0) %>% 
  mutate(phylum = factor(phylum, levels = tree$tip.label)) %>% 
  column_to_rownames('phylum')
```


```{r}
tree <- ggtree(tree, layout = 'circular', branch.length = 'none')

p.tree <- tree +
  geom_tiplab2(aes(text = label), size = 2, align = F, offset = 8)

p.tree
```


```{r}
gheatmap(p.tree, scores, 
         width = 0.75,
         colnames_angle = 285, hjust = 0, colnames_position = 'bottom', colnames_offset_y = 0.4,
         low = '#F0F1EB', high = '#444431',
         font.size = 6/.pt,
         legend_title = 'Average module completeness score'
         ) +
  theme(
    legend.title = element_text(size = 6),
    legend.text = element_text(size = 6),
    legend.position = 'bottom'
  )
```
```{r}
ggsave(filename = 'figures/heatmap.png' , width = 180, height = 180, units = 'mm')

# Number of genomes after each phylum
# Instead of average completeness score, count the number of genomes with the complete (or near-complete) module
```


# Bacteria
## Bacterial phylogeny - phylum


```{r}
#| label: bacterial-representatives

# Read the tree (bacterial) tree and remove the 'GB' or 'RS' prefix
tree <- ape::read.tree('../gtdb/bac120_r232.tree')
tree$tip.label <- gsub('^.._', '', tree$tip.label)

# getMRCA() only works when two or more tips are provided. So remove clades with a single representative

i <- gtdb %>%
  filter(domain == 'Bacteria') %>% 
  select(accession, phylum) %>% 
  # Consider only genomes represented in the tree
  filter(accession %in% tree$tip.label) %>%
  tibble::deframe()

# Assigns the accessions of each phyla in a list
groups <- split(names(i), i)

# Compute one representative node per phylum
phy_nodes <- sapply(groups, function(tips) {
  if (length(tips) > 1) {
    ape::getMRCA(tree, tips) # If the clade has more than one genome, find the MRCA
  } else {
    which(tree$tip.label == tips) # If only one tip, return its position in the tree
  }
})
```



```{r}
#| label: bacterial-nodes

# Ensure node indices are numeric and named by phylum (if all is well, this shouldn't change the data)
phy_nodes <- as.numeric(phy_nodes)
names(phy_nodes) <- names(groups)

#node <- phy_nodes[[fam]]
#tree$node.label <- rep(NA, tree$Nnode) # Remove node labels and replace with NA

# Indices for tip labels
tip_idx <- seq_along(tree$tip.label)

# Assign phylum names either to tips or internal nodes
for (p in names(phy_nodes)) {
  node <- phy_nodes[[p]]
  
  if (node <= length(tree$tip.label)) {
    # If node index corresponds to a tip - rename the tip
    tree$tip.label[node] <- p
  } else {
    # Otherwise it's an internal node → assign node label
    tree$node.label[node - length(tree$tip.label)] <- p
  }
}
```


```{r}
#| label: collapse-bacterial-tree


# Plot and collapse ('bactree' for 'collapsed tree')
bactree <- ggtree(tree)

# Collapse clades (multi-tip phyla) into single branches
for (p in names(groups)) {
  tips <- groups[[p]]
  
  if (length(tips) > 1) {
    node <- phy_nodes[p]
    if (!is.null(node)) {
      bactree <- collapse(bactree, node = node)
    }
  }
}
```


```{r}
#| label: plot-bacteria

bactree +
  geom_tiplab(size = 2) +
  geom_nodelab(aes(label = label), size = 2, hjust = 0)

```


```{r}
#| label: export-bacterial-phyla

ggsave('figures/gtdb-r232-phylum-bac.png', height = 240, width = 180, units = 'mm')
```

