---
title: "Carbon fixation"
author: "george.westmeijer@umu.se"
date: now
format: html
toc: true
number_sections: true
theme: sandstone
mainfont: Helvetica
monofont: AndaleMono
fontsize: "8"
dpi: 300
highlight: textmate
df_print: paged
code-fold: true
code-summary: "Expand code"
self_contained: true
embed-resources: true
keep_md: false
encoding: "UTF-8"
---

```{r}
#| label: setup
#| include: false
#| echo: false

knitr::opts_chunk$set(cache = TRUE)
```


```{r}
#| label: libraries
#| cache: false
#| include: false

library(dplyr)
library(tidyr)
library(tibble)
library(forcats)
library(ggplot2)
library(readr)
library(patchwork)
library(grateful)
library(ape)
library(ggtree)
library(mclink)
```

# Read input files

The file 'c-fixation-kos.gtdb-hits.tsv.gz' contains the genomes' accessions that have one or more hits of KOs part of a carbon fixation module.  


```{r}
#| label: read-taxonomy

gtdb <- read_tsv('../gtdb/ar53_metadata_r232.tsv.gz', show_col_types = F) %>%
  select(accession, ncbi_genbank_assembly_accession, taxonomy = gtdb_taxonomy, gtdb_representative) %>%
  mutate(accession = gsub('^.._', '', accession)) %>%
  mutate(taxonomy = gsub('.__', '', taxonomy))

gtdb <- read_tsv('../gtdb/bac120_metadata_r232.tsv.gz', show_col_types = F) %>%
  select(accession, ncbi_genbank_assembly_accession, taxonomy = gtdb_taxonomy, gtdb_representative) %>%
  mutate(accession = gsub('^.._', '', accession)) %>%
  mutate(taxonomy = gsub('.__', '', taxonomy)) %>%
  rbind(gtdb) %>% 
  pivot_longer(1:2, values_to = 'accession') %>% 
  select(accession, taxonomy, gtdb_representative) %>% 
  separate(taxonomy, into = c('domain', 'phylum', 'class', 'order', 'family', 'genus', 'species'), sep = ';') %>% 
  distinct() %>% 
  # Keep only species representatives
  filter(gtdb_representative) %>% 
  group_by(phylum) %>% mutate(n.rep = n()) %>% ungroup() 
```


```{r}
kf <- list.files('data/kofam', full.names = T, pattern = '^GCA')
ref <- markers$ko %>% unique()


kofam <- tibble()
for (file in kf) {
  kofam <- read_tsv(file, col_types = 'cccdddc') %>% 
    filter(ko %in% ref & !is.na(sign)) %>%
    mutate(accession = stringr::str_sub(file, start = 12, end = 26)) %>% 
    select(accession, ko, score, evalue) %>% 
    rbind(kofam)
}
```



```{r}
#| label: read-gtdb-phyla

gtdb <- read_tsv('../gtdb/gtdb_phyla_r232.tsv.gz', col_types = 'ccc') %>% 
  group_by(phylum) %>% 
  mutate(n = n()) %>% ungroup()
```

The KOfam output has been filtered by threshold (*), thereby only retaining KOs with a score above this threshold. In addition, some genome accessions / orthologous genes had multiple combinations, this was solved by running slice_max() and slice_min() in combination with group_by() for the threshold score and the e-value, respectively. 


```{r}
#| label: read-kofam

hits <- read_tsv('data/kofam-besthits.tsv.gz', col_types = 'ccdcdd') %>% 
  select(-score, -evalue) %>% 
  distinct()

scores <- read_tsv('data/mclink-scores.tsv.gz', col_types = 'ccd') %>% 
  left_join(gtdb, by = 'accession')

markers <- read_tsv('data/kos-markers.tsv', col_types = 'cccccclclcc') %>% 
  select(ko, module, marker, accession, model_organism) %>% 
  filter(marker) %>% 
  mutate(module = gsub('https://www.kegg.jp/module/', '', module))
  

# GCF_000010065.1 (S. elongatus)  M00165 Complete (2)
# GCF_000017945.1 (I. hospitalis) M00374 Complete (1)
# GCF_000018865.1 (C. aurantiacus) M00376 Complete (4)
# GCF_000247605.1 (A. woodii) M00377 Complete (1)
# GCF_000016605.1 (M. sedula) M00375 Complete (1)
# GCF_000006985.1 (C. tepidum) M00173 Not complete (3/8)
```


```{r}
#| label: read-mclink

scores <- read_tsv('data/mclink-scores.tsv.gz', col_types = 'ccd')

kos <- read_tsv('data/mclink-kos.tsv.gz', col_types = 'ccc') %>% 
  filter(ko %in% markers$ko) %>% 
  # M00173 had two enzymes as markers that each contain four subunits, thus eight KOs in total. 
  # To check for completeness, the subunits of each enzyme are grouped and two different subunits are required
  # These KOs are not part of any of the other modules
  mutate(ko = case_when(
    ko == 'K00170' ~ 'K00169',
    ko == 'K00171' ~ 'K00169',
    ko == 'K00172' ~ 'K00169',
    ko == 'K00175' ~ 'K00174',
    ko == 'K00176' ~ 'K00174',
    ko == 'K00177' ~ 'K00174',
    .default = ko
  )) %>% 
  distinct() %>% 
  # Count the number of marker genes for each module/accession id combination
  group_by(module, accession) %>% summarise(n = n(), .groups = 'drop') %>% 
  mutate(complete = case_when(
    module == 'M00165' ~ TRUE,
    module == 'M00377' ~ TRUE,
    module == 'M00173' & n == 2 ~ TRUE,
    module == 'M00376' ~ TRUE,
    module == 'M00375' ~ TRUE,
    module == 'M00374' ~ TRUE,
    .default = FALSE
  )) %>% 
  filter(complete) %>%
  # These genomes encode for all the marker genes for the respective pathway
  select(accession, module)
```



```{r}
#| label: mclink

# M00165 (CBB) has 12 blocks
# M00173 (Reverse citrate) has <nightmare>
# M00374 (DC-HB cycle) has 21 blocks
# M00375 (HP/HB cycle) has 17 blocks
# M00376 (3-HP bi-cycle) has 16 blocks (15 in the case of K01847)
# M00377 (WL pathway) has 14 blocks (13 in the case of K05299 + K15022)

# This is the actual analysis
mcl <- KO_pathway_ref %>% 
  filter(Module_Entry %in% c('M00165','M00377','M00173','M00376','M00375','M00374', 'M00620', 'M00579')) %>% 
  mclink(data = data.frame(
    hits %>%
      mutate(presence = 1) %>% 
      pivot_wider(names_from = 'accession', values_from = 'presence', values_fill = 0)), 
    ref = .)
```


```{r}
#| label: mclink-extract

# This extracts the data
scores <- mcl[['coverage']] %>% 
  as_tibble() %>% 
  rename(module = Module_Entry) %>%
  select(-Level_2, -Level_3, -Definition, -Module_Name) %>% 
  pivot_longer(2:ncol(.), names_to = 'accession', values_to = 'score')

kos <- mcl[['detected_KOs']] %>% 
  as_tibble() %>% 
  rename(module = Module_Entry) %>%
  select(-Level_2, -Level_3, -Definition, -Module_Name) %>% 
  pivot_longer(2:ncol(.), names_to = 'accession', values_to = 'ko') %>% 
  # Separate the column with the KO id's
  separate_longer_delim(ko, delim = ' ') %>% 
  # Only consider marker genes
  filter(ko %in% markers$ko) %>% 
  # M00173 had two enzymes as markers that each contain four subunits, thus eight KOs in total. 
  # To check for completeness, the subunits of each enzyme are grouped and two different subunits are required
  # These KOs are not part of any of the other modules
  mutate(ko = case_when(
    ko == 'K00170' ~ 'K00169',
    ko == 'K00171' ~ 'K00169',
    ko == 'K00172' ~ 'K00169',
    ko == 'K00175' ~ 'K00174',
    ko == 'K00176' ~ 'K00174',
    ko == 'K00177' ~ 'K00174',
    .default = ko
  )) %>% 
  distinct() %>% 
  # Count the number of marker genes for each module/accession id combination
  group_by(module, accession) %>% summarise(n = n(), .groups = 'drop') %>% 
  mutate(complete = case_when(
    module == 'M00165' ~ TRUE,
    module == 'M00377' ~ TRUE,
    module == 'M00173' & n == 2 ~ TRUE,
    module == 'M00376' ~ TRUE,
    module == 'M00375' ~ TRUE,
    module == 'M00374' ~ TRUE,
    .default = FALSE
  )) %>% 
  filter(complete) %>%
  # These genomes encode for all the marker genes for the respective pathway
  select(accession, module)
```


```{r}
#| label: score-cutoff

cutoff <- scores %>% 
  # inner_join() implies that only select genomes that encode for the marker genes
  inner_join(kos, by = c('accession', 'module')) %>% 
  # If the module score < 0.75, count it as complete
  mutate(score = if_else(score > 0.75, 1, 0)) %>% 
  inner_join(gtdb, by = 'accession') %>% 
  # Summarize the scores for each phylum/module combination
  group_by(phylum, module) %>% 
  summarise(n.marker = n(), n.complete = sum(score), .groups = 'drop') %>%
  # Some phyla are not represented but should get a value
  right_join(gtdb %>% select(phylum, n) %>% distinct(), by = 'phylum', relationship = 'many-to-one') %>% 
  mutate(complete = round(n.marker / n * 100, digits = 1))
```


```{r}
#| label: read-representatives

i <- list.files('data/c-fixation-genomes/kofam_scan', full.names = T)

kofam <- tibble()
for (file in i) {
  kofam <- read_tsv(file, col_types = 'cccdddc') %>% 
    mutate(accession = stringr::str_extract(file, pattern = 'GCF_000[0-9]*5.1')) %>% 
    select(accession, orf = accno, ko, thrshld, score, evalue) %>%
    inner_join(markers %>% select(ko, module), by = 'ko', relationship = 'many-to-many') %>% 
    rbind(kofam)
}

# GCF_000010065.1 (S. elongatus)  M00165 Complete (2)
# GCF_000017945.1 (I. hospitalis) M00374 Complete (1)
# GCF_000018865.1 (C. aurantiacus) M00376 Complete (4)
# GCF_000247605.1 (A. woodii) M00377 Complete (1)
# GCF_000016605.1 (M. sedula) M00375 Complete (1)
# GCF_000006985.1 (C. tepidum) M00173 Not complete (3/8)
```



```{r}
#| label: model-organisms-check

hits %>% 
  inner_join(markers, by = c('ko', 'accession'), relationship = 'many-to-many') %>%
  group_by(model_organism, module) %>% 
  summarise(n = n_distinct(ko), .groups = 'drop')
```



# KO matrix

The mclink function is used to calculate the completeness of the six relevant modules (M00165, M00377, M00173, M00376, M00375, M00374). The presence of any KO for each step indicates completeness, and the completeness is a score ranging from 0 to 1. The average value of this score is the genome-specific module completeness score. For example, a score of 0.9 for a module containing 10 steps (reactions) implies 9/10 genes were present.  


# Archaea
## Collapse archaeal tree with single representative

```{r}
#| label: drop-nwk

# There are 25 archaeal phyla in GTDB 232

tree <- ape::read.tree('../gtdb/ar53_r232.nwk')
tree$tip.label <- gsub('^.._', '', tree$tip.label) # 25 archaeal phyla in the tree

# Select a single representative per phylum, 25 in total
rep <- gtdb %>%
  filter(domain == 'Archaea') %>% 
  group_by(phylum) %>%
  slice(1) %>% pull(accession)

# Drop the genomes not among the representatives
tree <- ape::keep.tip(tree, rep)

# Decorate the tip labels with phyla names

# Export the tree
ape::write.tree(tree, file = 'data/arc_phyla_226.nwk')
```



```{r}
#| label: tiplabels

i <- tree$tip.label %>% 
  tibble() %>% 
  rename(accession = 1) %>% 
  left_join(gtdb, by = 'accession') %>% 
  select(phylum) %>% 
  left_join(cutoff, by = 'phylum', relationship = 'one-to-many') %>% 
  group_by(phylum) %>% slice_max(n.complete, n = 1) %>% ungroup() %>% 
  replace_na(list('n.complete' = 0)) %>% 
  mutate(tip.lab = if_else(
    is.na(module), 
    paste(phylum, ' (', n.complete, '/', n, ')', sep = ''),
    paste(phylum, ' (', n.complete, '/', n, ', ', module, ')', sep = '')
    )) %>% 
  select(phylum, tip.lab)
```




```{r}
#| label: hits-score

scores <- tree$tip.label %>% 
  tibble() %>% select(accession = 1) %>%
  left_join(gtdb, by = 'accession') %>% select(accession, phylum) %>% 
  left_join(cutoff, by = 'phylum', relationship = 'one-to-many') %>% 
  replace_na(list('module' = 'M00165', 'n.marker' = 0)) %>% 
  # Some phyla are not among the 'hits', thus have a score of 0
  select(-n.complete, -n) %>%
  pivot_wider(names_from = 'module', values_from = 'complete', values_fill = 0) %>%
  inner_join(i, by = 'phylum') %>% 
  #mutate(phylum = factor(phylum, levels = tree$tip.label)) %>%
  column_to_rownames('tip.lab') %>% 
  select(-accession, -phylum)
```


```{r}
#| label: archaea-ggtree

tree$tip.label <- i$tip.lab

tree <- ggtree(tree, layout = 'circular', branch.length = 'none')
```


```{r}
#| label: plot-archaeatree

p.tree <- tree +
  geom_tiplab2(aes(text = label), size = 2, align = F, offset = 5)

p.tree
```


```{r}
#| label: archaeal-heatmap

gheatmap(p.tree, scores, 
         width = 0.4,
         colnames_angle = 285, hjust = 0, colnames_position = 'bottom', colnames_offset_y = 0.4,
         low = '#f0f3f3', high = '#046C9A',
         font.size = 6/.pt,
         legend_title = 'Genomes with complete module (%)'
         ) +
  theme(
    legend.title = element_text(size = 6),
    legend.text = element_text(size = 6),
    legend.position = 'bottom',
    legend.margin = margin(t = 1.1, 0, 0, 0, 'cm'),
    legend.background = element_rect(fill = 'transparent'),
    plot.margin = margin(1.8, 1, 1, 1, 'cm')
  )
```
```{r}
ggsave(filename = 'figures/heatmap.png' , width = 120, height = 120, units = 'mm')

# Number of genomes after each phylum
# Instead of average completeness score, count the number of genomes with the complete (or near-complete) module
```


# Bacteria
## Bacterial phylogeny - phylum


```{r}
#| label: drop-nwk-bacteria

# There are 25 archaeal phyla in GTDB 232

tree <- ape::read.tree('../gtdb/bac120_r232.tree')
tree$tip.label <- gsub('^.._', '', tree$tip.label) # 25 archaeal phyla in the tree

# Select a single representative per phylum, 25 in total
rep <- gtdb %>%
  filter(domain == 'Bacteria') %>% 
  group_by(phylum) %>%
  slice(1) %>% pull(accession)

# Drop the genomes not among the representatives
tree <- ape::keep.tip(tree, rep)

# Decorate the tip labels with phyla names

# Export the tree
ape::write.tree(tree, file = 'data/bac_phyla_232.nwk')
```


```{r}
#| label: tiplabels

i <- tree$tip.label %>% 
  tibble() %>% 
  rename(accession = 1) %>% 
  left_join(gtdb, by = 'accession') %>% 
  select(phylum) %>% 
  left_join(cutoff, by = 'phylum', relationship = 'one-to-many') %>% 
  group_by(phylum) %>% slice_max(n.complete, n = 1, with_ties = FALSE) %>% ungroup() %>% 
  replace_na(list('n.complete' = 0)) %>% 
  mutate(tip.lab = if_else(
    is.na(module), 
    paste(phylum, ' (', n.complete, '/', n, ')', sep = ''),
    paste(phylum, ' (', n.complete, '/', n, ', ', module, ')', sep = '')
    )) %>% 
  select(phylum, tip.lab)
```


```{r}
#| label: bacteria-score

scores <- tree$tip.label %>% 
  tibble() %>% select(accession = 1) %>%
  left_join(gtdb, by = 'accession') %>% select(accession, phylum) %>% 
  left_join(cutoff, by = 'phylum', relationship = 'one-to-many') %>% 
  replace_na(list('module' = 'M00165', 'complete' = 0)) %>% 
  # Some phyla are not among the 'hits', thus have a score of 0
  select(-n.complete, -n) %>%
  pivot_wider(names_from = 'module', values_from = 'complete', values_fill = 0) %>%
  inner_join(i, by = 'phylum') %>% 
  #mutate(phylum = factor(phylum, levels = tree$tip.label)) %>%
  column_to_rownames('tip.lab') %>% 
  select(-accession, -phylum)
```



```{r}
#| label: bacteria-ggtree

tree$tip.label <- i$tip.lab

tree <- ggtree(tree, layout = 'circular', branch.length = 'none')
```


```{r}
#| label: plot-archaeatree

p.tree <- tree +
  geom_tiplab2(aes(text = label), size = 2, align = F, offset = 10)

p.tree
```


```{r}
#| label: archaeal-heatmap

gheatmap(p.tree, scores, 
         width = 0.4,
         colnames_angle = 285, hjust = 0, colnames_position = 'bottom', colnames_offset_y = 0.4,
         low = '#f0f3f3', high = '#046C9A',
         font.size = 6/.pt,
         legend_title = 'Genomes with complete module (%)'
         ) +
  theme(
    legend.title = element_text(size = 6),
    legend.text = element_text(size = 6),
    legend.position = 'bottom',
    legend.margin = margin(t = 1.1, 0, 0, 0, 'cm'),
    legend.background = element_rect(fill = 'transparent'),
    plot.margin = margin(1.8, 1, 1, 1, 'cm')
  )
```
```{r}
ggsave(filename = 'figures/heatmap_bacteria.pdf' , width = 240, height = 240, units = 'mm')

# Number of genomes after each phylum
# Instead of average completeness score, count the number of genomes with the complete (or near-complete) module
```

